\name{}: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-04-10 paper/main.tex
.

In order to design representations concerning both top-down and bottom-up features for proteins and drugs,respectively, we utilize protein--protein interaction (PPI) networks which have shown great results in protein function prediction \cite{Vazquez2003} and antiviral drug--target discovery \cite{Ackerman2019}. 
However, only few applied this context to the task of DTI prediction in humans. The method of graph convolutional neural networks has been applied successfully to PPI networks \cite{Zitnik2017}, granting valuable information for medical diagnostics, closely related to phenotypic information.

Only recently, the technique of graph learning was introduced by
\citet{GCNConv} through graph convolution algorithms, and improved and
altered under usage of different kernels \cite{ChebConv, ARMAConv},
attention mechanisms \cite{GATConv}, random walks \cite{APPNPConv},
and mixtures of both \cite{SAGEConv}. While based on diverse systems,
they can be relevant for testing distinct hypothesis for given
graphs. While convolutional filters are suitable for finding patterns
among the given graph, attention mechanisms are more relevant for
discovery of important regions within. Lately, graph learning
approaches found application for computing compound representations
for DTI prediction.

Eventually, we further address potential biases and skews within DTI prediction datasets. \todo[inline]{Introduction}
Within DTI prediction, there are potential biases resulting from the
underlying datasets \citep{Pahikkala2014}. First, novel drugs are
often designed by altering non-functional components of a drug,
leading to two and more very similar drugs designed to target the same
proteins \cite{Overington2006}. This can result in a bias when it
leads to hidden duplicates that can distribute among the train/test
split, resulting in a better (measured) predictive performance than
would be expected when the model is applied to identify drugs that
target a protein for which no drugs yet exist. Second, some proteins
(which we call \textit{hub proteins}) have significantly more known
interactions with drugs than others. In the STITCH database, $5\%$ of
the proteins have $40\%$ of the interactions, and similar
distributions are present in the Yamanishi and Drugbank
\cite{Drugbank2007, Drugbank2017} datasets; preferentially predicting
these proteins may increase predictive performance while not
reflecting the actual performance when applied to a new protein (e.g.,
a protein for which no interactions are known). These differences in
the number of drugs targeting certain proteins may be the result of
study bias where more ``valuable'' proteins have more drugs designed
to target them due to their involvement in more common diseases (or
diseases for which drugs can be more profitably marketed). This may affect common \cite{Survey2018} cross-validation splitting schemes, that allow exploitation of these biases within DTI prediction.

>>>>

Progress in machine learning using graph neural networks can allow us
to test this hypothesis and combine both bottom-up and top-down
features with a network in a single machine learning model. 
In particular, Graph Convolutional Networks \cite{GCNConv} and
their variants using different types of kernels \cite{ChebConv,
  ARMAConv}, including attention mechanisms \cite{GATConv}, and
different forms of exploring node neighborhoods \cite{APPNPConv,
  SAGEConv} can combine different types of features and graph-based
information. % While based on diverse systems,
% they can be relevant for testing distinct hypothesis for given
% graphs. While convolutional filters are suitable for finding patterns
% among the given graph, attention mechanisms are more relevant for
% discovery of important regions within. Lately, graph learning
% approaches found application for computing compound representations
% for DTI prediction.
They have previously been applied for a number of tasks, including
prediction of protein functions \cite{Zitnik2017}, \todo{add 2 more,
  and then one on DTI prediction}.

One challenge in DTI prediction are potential biases resulting from
the underlying datasets \citep{Pahikkala2014} which may affect model
evaluation and comparison. First, novel drugs are often designed by
altering non-functional components of a drug, leading to two and more
very similar drugs designed to target the same proteins
\cite{Overington2006}. This can result in a bias when it leads to
hidden duplicates or highly similar compounds that are distributed
among training and evaluation dataset, resulting in a better
(measured) predictive performance than would be expected when the
model is applied to identify drugs that target a protein for which no
drugs yet exist. Second, some proteins (which we call \textit{hub
  proteins}) have significantly more known interactions with drugs
than others. In the STITCH database, $5\%$ of the proteins have $40\%$
of the interactions, and similar distributions are present in other
datasets \cite{Drugbank2007, Drugbank2017}; preferentially predicting
these proteins may increase predictive performance while not
reflecting the actual performance when applied to a new protein (i.e.,
a protein for which no interactions are known). These differences in
the number of drugs targeting certain proteins may be the result of
study bias where more ``valuable'' proteins have more drugs designed
to target them due to their involvement in more common diseases (or
diseases for which drugs can be more profitably marketed). This may
affect common evaluation schemes where it is possible
to exploit these biases within DTI prediction \cite{Survey2018}.

% In order to design representations concerning both top-down and bottom-up features for proteins and drugs,respectively, we utilize protein--protein interaction (PPI) networks which have shown great results in protein function prediction \cite{Vazquez2003} and antiviral drug--target discovery \cite{Ackerman2019}. 
% However, only few applied this context to the task of DTI prediction in humans.
